Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-07-12 article-models-ttinternal.tex
a bit of progress on cftsubst

\subsubsection{Currying for weakening}
Let $\cftalg{A}$ be a CFT-algebra in context $\Gamma$ and let
\begin{equation*}
\jhom
  {{{\Gamma}{\cftalgc{\cftalg{A}}}}{\cftalgf{\cftalg{A}}}}
  {\ctxwk{\cftalgf{\cftalg{A}}}{\cftfamalg{\cftalg{A}}}}
  {\cftfamalg{\cftfamalg{\cftalg{A}}}}
  {\cftwk{\cftalg{A}}}
\end{equation*}
be a CFT-homomorphism. The abstraction of the rules in \autoref{comp-ew} is the
condition that the diagram
\begin{equation}\label{diag:cftwk-curry}
\begin{tikzcd}[column sep=large]
\ctxwk
  {\ctxext{\cftalgf{\cftalg{A}}}{\cftalgf{\cftfamalg{\cftalg{A}}}}}
  {\cftfamalg{\cftalg{A}}}
  \ar{r}{\ctxwk{\cftalgf{\cftfamalg{\cftalg{A}}}}{\cftwk{\cftalg{A}}}}
  \ar{dr}[swap]{\jcomp{}{\cftfamext}{\cftwk{\cftalg{A}}}}
& \ctxwk{\cftalgf{\cftfamalg{\cftalg{A}}}}{\cftfamalg{\cftfamalg{\cftalg{A}}}}
  \ar{d}{\jcomp{}{\cftctxext}{\cftwk{\cftalg{A}}}}
  \\
& \cftfamalg{\cftfamalg{\cftfamalg{\cftalg{A}}}}
\end{tikzcd}
\end{equation}
of CFT-algebras in context 
$\ctxext
    {{{\Gamma}{\cftalgc{\cftalg{A}}}}{\cftalgf{\cftalg{A}}}}
    {\cftalgf{\cftfamalg{\cftalg{A}}}}
$ %
commutes.

\begin{lem}
Let $\cftalg{A}$ be a CFT-algebra in context gamma with a CFT-homomorphism 
$ \jhom
  {{{\Gamma}{\cftalgc{\cftalg{A}}}}{\cftalgf{\cftalg{A}}}}
  {\ctxwk{\cftalgf{\cftalg{A}}}{\cftfamalg{\cftalg{A}}}}
  {\jcomp{}{\cftctxext}{\cftfamalg{\cftalg{A}}}}
  {\cftwk{\cftalg{A}}}
$
satisfying the judgmental equality of \autoref{diag:cftwk-curry}. Then we
have the judgmental equality
\begin{equation*}
\apply\cftwk{\cftalg{A}}{\gamma}{\apply\cftfamext{\gamma}{a}{p}}
\jdeq
\jcomp{}
    {\apply\cftwk{\cftalg{A}}{\gamma}{a}}
    {\apply\cftwk{\cftalg{A}}{\apply\cftctxext{\gamma}{a}}{p}}
\end{equation*}
of CFT-homomorphisms from $\subst{\gamma}{\cftfamalg{\cftalg{A}}}$ to 
$\subst{p}{{a}{{\gamma}{\cftalg{F_{F_{F_A}}}}}}$ in context $\Gamma$,
for every $\gamma:A$, $a:\subst{\gamma}{\cftalgf{\cftalg{A}}}$ and
$p:\subst{a}{{\gamma}{\cftalgf{\cftfamalg{\cftalg{A}}}}}$. 
\end{lem}

@@

\begin{defn}
A weakening algebra in context $\Gamma$ is a CFT-algebra $\cftalg{A}$ in context
$\Gamma$ for which there is a CFT-homomorphism
\begin{equation*}
\jhom
  {{{\Gamma}{\cftalgc{\cftalg{A}}}}{\cftalgf{\cftalg{A}}}}
  {\ctxwk{\cftalgf{\cftalg{A}}}{\cftfamalg{\cftalg{A}}}}
  {\cftalg{F_{F_A}}}
  {\cftwk{\cftalg{A}}}
\end{equation*}
satisfying the judgmental equalities of %
\autoref{diag:cftwk-cftwk,eq:cftwk-cftextcurry,eq:cftwk-cftempcurry}.
\end{defn}

@@

\begin{proof}
To show that $\cftwk{\cftfamalg{\cftalg{A}}}$ satisfies the Currying rule, we
must verify that the diagram
\begin{equation*}
\begin{tikzcd}[column sep=large]
\ctxwk
  {\ctxext{\cftalgf{\cftfamalg{\cftalg{A}}}}{\cftalgf{\cftfamalg{\cftfamalg{\cftalg{A}}}}}}
  {\cftfamalg{\cftfamalg{\cftalg{A}}}}
  \ar{r}{\ctxwk{\cftalgf{\cftfamalg{\cftfamalg{\cftalg{A}}}}}{\cftwk{\cftfamalg{\cftalg{A}}}}}
  \ar{dr}[swap]{\jcomp{}{\jcomp{}{\cftctxext}{\cftfamext}}{\cftwk{\cftfamalg{\cftalg{A}}}}}
& \ctxwk{\cftalgf{\cftfamalg{\cftfamalg{\cftalg{A}}}}}{\cftfamalg{\cftfamalg{\cftfamalg{\cftalg{A}}}}}
  \ar{d}{\jcomp{}{\cftfamext}{\cftwk{\cftfamalg{\cftalg{A}}}}}
  \\
& \cftfamalg{\cftfamalg{\cftfamalg{\cftfamalg{\cftalg{A}}}}}
\end{tikzcd}
\end{equation*}
commutes judgmentally. We do this by showing that each of the ingredients of this
diagram is judgmentally equal to the corresponding ingredient of the diagram in
\autoref{diag:cftwk-curry} pulled back along $\cftctxext$. Thus, we are going
to show that
\begin{align*}
\ctxwk{\cftalgf{\cftfamalg{\cftfamalg{\cftalg{A}}}}}{\cftwk{\cftfamalg{\cftalg{A}}}}
& \jdeq
\jcomp{}{\cftctxext}{\ctxwk{\cftalgf{\cftfamalg{\cftalg{A}}}}{\cftwk{\cftalg{A}}}}
  \tag{1}
  \\
\jcomp{}{\cftfamext}{\cftwk{\cftfamalg{\cftalg{A}}}}
& \jdeq
\jcomp{}{\cftctxext}{\jcomp{}{\cftctxext}{\cftwk{\cftalg{A}}}}
  \tag{2}
  \\
\jcomp{}{\jcomp{}{\cftctxext}{\cftfamext}}{\cftwk{\cftfamalg{\cftalg{A}}}}
& \jdeq
\jcomp{}{\cftctxext}{\jcomp{}{\cftfamext}{\cftwk{\cftalg{A}}}}
  \tag{3}
\end{align*}
For the first, we have the judgmental equalities

@@

For the forth, note that we have the judgmental equalities

@@

Thus we may conclude that the diagram above of which we had the aim to show that
it commutes judgmentally is a pullback of the diagram in \autoref{diag:cftwk-cftwk}

@@

@@

\begin{defn}
Let $\cftalg{A}$ be a CFT-algebra in context $\Gamma$ and let
\begin{equation*}
\jhom
  {{{{\Gamma}{\cftalgc{\cftalg{A}}}}{\cftalgf{\cftalg{A}}}}{\cftalgt{\cftalg{A}}}}
  {\ctxwk{\cftalgt{\cftalg{A}}}{\cftalg{F_{F_A}}}}
  {\cftfamalg{\cftalg{A}}}
  {\cftsubst{\cftalg{A}}}
\end{equation*}
be a CFT-homomorphism. We say that the pair $(\cftalg{A},\cftsubst{\cftalg{A}})$
is a substitution algebra in context $\Gamma$ if the diagram
\begin{equation}\label{diag:substalg}
\begin{tikzcd}
X
  \ar{r}
  \ar{d}
& X
  \ar{d}
  \\
X 
  \ar{r}
& X
\end{tikzcd}
\end{equation}
of CFT-algebras in context ... commutes judgmentally.
\end{defn}

>>>>

\subsubsection{Currying for weakening}
Let $\cftalg{A}$ be a CFT-algebra in context $\Gamma$ and let
\begin{equation*}
\jhom
  {{{\Gamma}{\cftalgc{\cftalg{A}}}}{\cftalgf{\cftalg{A}}}}
  {\ctxwk{\cftalgf{\cftalg{A}}}{\cftfamalg{\cftalg{A}}}}
  {\cftfamalg{\cftfamalg{\cftalg{A}}}}
  {\cftwk{\cftalg{A}}}
\end{equation*}
be a CFT-homomorphism. The abstraction of the rules in \autoref{comp-ew} is the
condition that the diagram
\begin{equation}\label{diag:cftwk-cftextcurry}
\begin{tikzcd}[column sep=large]
\ctxwk
  {\ctxext{\cftalgf{\cftalg{A}}}{\cftalgf{\cftfamalg{\cftalg{A}}}}}
  {\cftfamalg{\cftalg{A}}}
  \ar{r}{\ctxwk{\cftalgf{\cftfamalg{\cftalg{A}}}}{\cftwk{\cftalg{A}}}}
  \ar{dr}[swap]{\jcomp{}{\cftfamext}{\cftwk{\cftalg{A}}}}
& \ctxwk{\cftalgf{\cftfamalg{\cftalg{A}}}}{\cftfamalg{\cftfamalg{\cftalg{A}}}}
  \ar{d}{\jcomp{}{\cftctxext}{\cftwk{\cftalg{A}}}}
  \\
& \cftfamalg{\cftfamalg{\cftfamalg{\cftalg{A}}}}
\end{tikzcd}
\end{equation}
of CFT-algebras in context 
$\ctxext
    {{{\Gamma}{\cftalgc{\cftalg{A}}}}{\cftalgf{\cftalg{A}}}}
    {\cftalgf{\cftfamalg{\cftalg{A}}}}
$ %
commutes.

\begin{lem}
Let $\cftalg{A}$ be a CFT-algebra in context gamma with a CFT-homomorphism 
$ \jhom
  {{{\Gamma}{\cftalgc{\cftalg{A}}}}{\cftalgf{\cftalg{A}}}}
  {\ctxwk{\cftalgf{\cftalg{A}}}{\cftfamalg{\cftalg{A}}}}
  {\jcomp{}{\cftctxext}{\cftfamalg{\cftalg{A}}}}
  {\cftwk{\cftalg{A}}}
$
satisfying the judgmental equality of \autoref{diag:cftwk-cftextcurry}. Then we
have the judgmental equality
\begin{equation*}
\apply\cftwk{\cftalg{A}}{\gamma}{\apply\cftfamext{\gamma}{a}{p}}
\jdeq
\jcomp{}
    {\apply\cftwk{\cftalg{A}}{\gamma}{a}}
    {\apply\cftwk{\cftalg{A}}{\apply\cftctxext{\gamma}{a}}{p}}
\end{equation*}
of CFT-homomorphisms from $\subst{\gamma}{\cftfamalg{\cftalg{A}}}$ to 
$\subst{p}{{a}{{\gamma}{\cftalg{F_{F_{F_A}}}}}}$ in context $\Gamma$,
for every $\gamma:A$, $a:\subst{\gamma}{\cftalgf{\cftalg{A}}}$ and
$p:\subst{a}{{\gamma}{\cftalgf{\cftfamalg{\cftalg{A}}}}}$. 
\end{lem}

@@

\begin{defn}
A weakening algebra in context $\Gamma$ is a CFT-algebra $\cftalg{A}$ in context
$\Gamma$ for which there is a CFT-homomorphism
\begin{equation*}
\jhom
  {{{\Gamma}{\cftalgc{\cftalg{A}}}}{\cftalgf{\cftalg{A}}}}
  {\ctxwk{\cftalgf{\cftalg{A}}}{\cftfamalg{\cftalg{A}}}}
  {\cftalg{F_{F_A}}}
  {\cftwk{\cftalg{A}}}
\end{equation*}
satisfying the judgmental equalities of %
\autoref{diag:cftwk-cftwk,diag:cftwk-cftextcurry,eq:cftwk-cftempcurry}.
\end{defn}

@@

\begin{proof}
To show that $\cftwk{\cftfamalg{\cftalg{A}}}$ satisfies the Currying rule, we
must verify that the diagram
\begin{equation*}
\begin{tikzcd}[column sep=large]
\ctxwk
  {\ctxext{\cftalgf{\cftfamalg{\cftalg{A}}}}{\cftalgf{\cftfamalg{\cftfamalg{\cftalg{A}}}}}}
  {\cftfamalg{\cftfamalg{\cftalg{A}}}}
  \ar{r}{\ctxwk{\cftalgf{\cftfamalg{\cftfamalg{\cftalg{A}}}}}{\cftwk{\cftfamalg{\cftalg{A}}}}}
  \ar{dr}[swap]{\jcomp{}{\jcomp{}{\cftctxext}{\cftfamext}}{\cftwk{\cftfamalg{\cftalg{A}}}}}
& \ctxwk{\cftalgf{\cftfamalg{\cftfamalg{\cftalg{A}}}}}{\cftfamalg{\cftfamalg{\cftfamalg{\cftalg{A}}}}}
  \ar{d}{\jcomp{}{\cftfamext}{\cftwk{\cftfamalg{\cftalg{A}}}}}
  \\
& \cftfamalg{\cftfamalg{\cftfamalg{\cftfamalg{\cftalg{A}}}}}
\end{tikzcd}
\end{equation*}
commutes judgmentally. We do this by showing that each of the ingredients of this
diagram is judgmentally equal to the corresponding ingredient of the diagram in
\autoref{diag:cftwk-cftextcurry} pulled back along $\cftctxext$. Thus, we are going
to show that
\begin{align*}
\ctxwk{\cftalgf{\cftfamalg{\cftfamalg{\cftalg{A}}}}}{\cftwk{\cftfamalg{\cftalg{A}}}}
& \jdeq
\jcomp{}{\cftctxext}{\ctxwk{\cftalgf{\cftfamalg{\cftalg{A}}}}{\cftwk{\cftalg{A}}}}
  \tag{1}
  \\
\jcomp{}{\cftfamext}{\cftwk{\cftfamalg{\cftalg{A}}}}
& \jdeq
\jcomp{}{\cftctxext}{\jcomp{}{\cftctxext}{\cftwk{\cftalg{A}}}}
  \tag{2}
  \\
\jcomp{}{\jcomp{}{\cftctxext}{\cftfamext}}{\cftwk{\cftfamalg{\cftalg{A}}}}
& \jdeq
\jcomp{}{\cftctxext}{\jcomp{}{\cftfamext}{\cftwk{\cftalg{A}}}}
  \tag{3}
\end{align*}
For the first, we have the judgmental equalities
\begin{align*}
\ctxwk{\cftalgf{\cftfamalg{\cftfamalg{\cftalg{A}}}}}{\cftwk{\cftfamalg{\cftalg{A}}}}
& \jdeq
\ctxwk
  {\jcomp{}{\cftctxext}{\cftalgf{\cftfamalg{\cftalg{A}}}}}
  {\jcomp{}{\cftctxext}{\cftwk{\cftalg{A}}}}
  \\
& \jdeq
\jcomp{}{\cftctxext}{\ctxwk{\cftalgf{\cftfamalg{\cftalg{A}}}}{\cftwk{\cftalg{A}}}}
\end{align*}
For the second, we have the judgmental equalities
\begin{align*}
\jcomp{}{\cftfamext}{\cftwk{\cftfamalg{\cftalg{A}}}}
& \jdeq
  \jcomp{}{\cftfamext}{\jcomp{}{\cftctxext}{\cftwk{\cftalg{A}}}}
  \\
& \jdeq
  \jcomp{}{\cftctxext}{\jcomp{}{\cftctxext}{\cftwk{\cftalg{A}}}}
\end{align*}
For the third, note that we have the judgmental equalities
\begin{align*}
\jcomp{}{\jcomp{}{\cftctxext}{\cftfamext}}{\cftwk{\cftfamalg{\cftalg{A}}}}
& \jdeq
  \jcomp{}{\jcomp{}{\cftctxext}{\cftfamext}}{\jcomp{}{\cftctxext}{\cftwk{\cftalg{A}}}}
  \\
& \jdeq
\jcomp{}{\cftctxext}{\jcomp{}{\cftfamext}{\cftwk{\cftalg{A}}}}
\end{align*}
Thus we may conclude that the diagram above a pullback of the diagram in 
\autoref{diag:cftwk-cftextcurry} by $\cftctxext$ and therefore it commutes as desired.

@@

For the fourth, note that we have the judgmental equalities

@@

Thus we may conclude that the diagram above a pullback of the diagram in \autoref{diag:cftwk-cftwk}

@@

\begin{thm}
Suppose $\cftalg{A}$ is a projection algebra in context $\Gamma$. Then
$\cftfamalg{\cftalg{A}}$ with the internal identity term
\begin{equation*}
\cftidtm{\cftfamalg{\cftalg{A}}}\defeq\jcomp{}{\cftctxext}{\cftidtm{\cftalg{A}}}
\end{equation*}
is a projection algebra in context $\ctxext{\Gamma}{\cftalgc{\cftalg{A}}}$.
\end{thm}

\begin{thm}
Suppose $\cftalg{Q}$ is a projection algebra in context $\ctxext{\Gamma}{B}$ and
let $\jfam{\Gamma}{A}$. Then $\ctxwk{A}{\cftalg{Q}}$ together with the
internal identity term
\begin{equation*}
\cftidtm{\ctxwk{A}{\cftalg{Q}}}\defeq \ctxwk{A}\cftidtm{\cftalg{Q}}
\end{equation*}
is a projection algebra in context $\ctxext{{\Gamma}{A}}{B}$.
\end{thm}

\begin{thm}
Suppose $\cftalg{Q}$ is a projection algebra in context $\ctxext{{\Gamma}{A}}{P}$
and let $\jterm{\Gamma}{A}{x}$. Then $\subst{x}{\cftalg{Q}}$ together with the
internal identity term
\begin{equation*}
\cftidtm{\subst{x}{\cftalg{Q}}} \defeq \subst{x}{\cftidtm{\cftalg{Q}}}
\end{equation*}
is a projection algebra in context $\ctxext{\Gamma}{\subst{x}{P}}$.
\end{thm}

@@

\begin{defn}
Let $\cftalg{A}$ be a CFT-algebra in context $\Gamma$ and let
\begin{equation*}
\jhom
  {{{{\Gamma}{\cftalgc{\cftalg{A}}}}{\cftalgf{\cftalg{A}}}}{\cftalgt{\cftalg{A}}}}
  {\ctxwk{\cftalgt{\cftalg{A}}}{\cftalg{F_{F_A}}}}
  {\cftfamalg{\cftalg{A}}}
  {\cftsubst{\cftalg{A}}}
\end{equation*}
be a CFT-homomorphism. We say that the pair $(\cftalg{A},\cftsubst{\cftalg{A}})$
is a substitution algebra in context $\Gamma$ if the diagram
\begin{equation}\label{diag:cftsubst-cftsubst}
\begin{tikzcd}[column sep=9em]
\ctxwk
  { \cftalgt{\cftalg{A}}}
  { { \cftalgt{\cftfamalg{\cftfamalg{\cftalg{A}}}}}
    { \cftfamalg{\cftfamalg{\cftfamalg{\cftfamalg{\cftalg{A}}}}}}
    }
  \ar{r}
    { \ctxwk
        { \cftalgt{\cftalg{A}}}
        { \jcomp{}
            { \cftctxext}
            { \jcomp{}
                { \cftctxext}
                { \cftsubst{\cftalg{A}}}
              }
          }
      }
  \ar{d}[swap]
    { \ctxwk
        { { \cftalgt{\cftalg{A}}}
          { \cftalgt{\cftfamalg{\cftfamalg{\cftalg{A}}}}}
          }
        { \jcomp{}{\ctxwk{\cftalgt{\cftalg{A}}}{\jcomp{}{\cftctxext}{\cftfamext}}}{\cftfamhom{\cftsubst{\cftalg{A}}}}}
      }
  & 
\ctxwk
  { \cftalgt{\cftalg{A}}}
  { \cftfamalg{\cftfamalg{\cftfamalg{\cftalg{A}}}}}
  \ar{d}{ \cftfamhom{\cftsubst{\cftalg{A}}}}
  \\
\ctxwk
  { { \cftalgt{\cftalg{A}}}
    { \cftalgt{\cftfamalg{\cftfamalg{\cftalg{A}}}}}
    }
  { \jcomp{}
      { \ctxwk{\cftalgt{\cftalg{A}}}{\jcomp{}{\cftctxext}{\cftfamext}}}
      { \jcomp{}
          {\cftsubstc{\cftalg{A}}}
          {\cftfamalg{\cftfamalg{\cftalg{A}}}}
        }
    }
  \ar{r}
& \jcomp{}{\cftsubstc{\cftalg{A}}}{\cftfamalg{\cftfamalg{\cftalg{A}}}}
\end{tikzcd}
\end{equation}
of CFT-algebras in context 
$ \ctxext
    { { { { { {\Gamma}
              {\cftalgc{\cftalg{A}}}
              }
            { \cftalgf{\cftalg{A}}}
            }
          { \cftalgt{\cftalg{A}}}
          }
        { \ctxwk{\cftalgt{\cftalg{A}}}{\cftalgf{\cftfamalg{\cftalg{A}}}}}
        }
      { \ctxwk{\cftalgt{\cftalg{A}}}{\cftalgf{\cftfamalg{\cftfamalg{\cftalg{A}}}}}}
      }
    { \ctxwk{\cftalgt{\cftalg{A}}}{\cftalgt{\cftfamalg{\cftfamalg{\cftalg{A}}}}}}$
commutes judgmentally.
\end{defn}

\begin{rmk}
We shall have to verify that the diagram in \autoref{diag:cftsubst-cftsubst} is
well-typed.
\end{rmk}
